---
title: "Pooled 3-Cohort Neuropsych: Class 2 Life-Course Plots"
author: "Minghui (Grace) Wang"
date: "`r Sys.Date()`"
output: html_document
---

<!--
Pipeline step: 03 (run order: 00 -> 01 -> 02 -> 03 -> 04 -> 05 -> 06)
Purpose: Generate Class 2 (follow-up-as-time) life-course trajectory plots with
         baseline-age-anchored fitted curves. Must run BEFORE 04 — saves a y-axis
         range table that 04 reads for aligned scaling across model classes.
Inputs:  - Class 2 fitted model RDS from 01 (../output/models/)
         - neuropsych_pooled3_model_ready_xprize_with_covariates.csv (via setup_analysis_sample.R)
Key steps: Generate fixed-effect predictions at baseline ages 30–85 by 5 over
           follow-up grid 0–20yr; compute curve-based y-axis ranges; plot with
           dual y-axis (score + follow-up) and viridis color for baseline age.
Core functions: make_curve_y_axis() (helpers.R), make_pooled_test_dictionary() (workbook_writers.R)
Outputs: - Class 2 life-course PDFs (../output/plots/pooled3_03_lifecourse_plots_class2_crude.../)
         - class2_y_axis_table RDS (../output/tables/ — used by 04)
Downstream: 04 reads the y-axis range table for aligned scaling

-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,   message = FALSE, warning = FALSE)

library(tidyverse)
library(lme4)
library(lmerTest)

source("R/_load_all.R")

output_suffix <- format(Sys.Date(), "%m%d%Y")

```

## 0. Overview
This script generates Class 2 life-course plots for the pooled 3-cohort neuropsychological models. It loads fitted model objects from RDS (saved by `01_fit_models.Rmd`) instead of refitting.

**Run this script before `neuropsych_pooled3cohorts_04_plot_lifecourse_curves_model1_crude.Rmd`.** The resulting y-axis limits are reused in 04 for the Class 1 plots.

Class 2 models change as a function of follow-up time. For plotting, predictions are generated from selected baseline ages and displayed against attained age.

## 0.1 Choose analysis sample

```{r}
SAMPLE <- "2visit"
```

```{r}
covariates <- NULL
source("R/setup_analysis_sample.R")
```

## 1. Setup and data import

### 1.1 Define output folders

```{r}
model_tag <- if (SAMPLE == "2visit") "crude_2visit" else "crude"
dir_tag   <- if (SAMPLE == "2visit") "_2visit" else ""

plot_dir <- paste0(
  "../output/plots/pooled3_03_lifecourse_plots_class2_crude", dir_tag
)

class2_lifecourse_dir <- file.path(
  plot_dir,
  "baseline_age_anchored_trajectories"
)

dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(class2_lifecourse_dir, recursive = TRUE, showWarnings = FALSE)
```

## 2. Load fitted Class 2 model objects

```{r}
model_rds_dir <- "../output/models"

class2_fits <- readRDS(
  file.path(
  model_rds_dir,
  paste0("pooled3_", model_tag, "_class2_fits.rds")
)
)

class2_fits_success <- class2_fits |>
  filter(fit_success)

cat(
  "Loaded", nrow(class2_fits_success),
  "successfully fitted Class 2 model strata\n"
)

class2_fits_success |>
  select(
    test,
    sex_label,
    n_obs,
    n_subjects,
    subject_random_slope_used
  ) |>
  knitr::kable()
```



## 3. Plotting inputs

### 3.1 Create test dictionary

```{r}
test_dictionary_sheet <- make_pooled_test_dictionary( # defined in R/workbook_writers.R
  model_ready_pool
)

test_dictionary_sheet |>
  knitr::kable()

```
### 3.2 Define plot settings


```{r}
plot_age_min <- 30
plot_age_max <- 90


cohort_shapes <- c(
  "Original" = 17,
  "Offspring" = 16,
  "Gen3" = 15
)

```


### 3.3 Prepare baseline-age × maximum-follow-up background layer

```{r}
baseline_followup_background <- model_ready_analysis |>
  drop_na(
    test,
    sex_label,
    idtype_label,
    baseline_age,
    t_followup,
    score
  ) |>
  group_by(
    test,
    sex_label,
    idtype_label,
    dbgap_subject_id,
    baseline_age
  ) |>
  summarise(
    max_followup = max(t_followup, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    sex_label = factor(
      sex_label,
      levels = c("Women", "Men")
    ),
    idtype_label = factor(
      idtype_label,
      levels = c("Original", "Offspring", "Gen3")
    )
  ) |>
  filter(
    baseline_age >= plot_age_min,
    baseline_age <= plot_age_max
  ) |>
  left_join(
    test_dictionary_sheet |>
      select(
        test,
        test_label,
        xprize_subdomain,
        range
      ),
    by = "test"
  )

baseline_followup_background |>
  count(
    test,
    sex_label,
    idtype_label
  ) |>
  arrange(
    test,
    sex_label,
    idtype_label
  ) |>
  knitr::kable()
```

### 3.4 Check observed baseline-age and follow-up support by test

```{r}

observed_score_range_by_test <- baseline_followup_background |>
  group_by(
    test,
    test_label,
    xprize_subdomain,
    sex_label
  ) |>
  summarise(
    n_subjects = n_distinct(dbgap_subject_id),
    min_baseline_age = min(baseline_age, na.rm = TRUE),
    max_baseline_age = max(baseline_age, na.rm = TRUE),
    min_max_followup = min(max_followup, na.rm = TRUE),
    max_max_followup = max(max_followup, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(
    xprize_subdomain,
    test,
    sex_label
  )

observed_score_range_by_test |>
  knitr::kable(digits = 2)
```


## 4. Class 2 · Follow-up as Time life-course curves

### 4.1 Define baseline-age anchors and follow-up grid

```{r}

baseline_age_anchors <- seq(
  30,
  85,
  by = 5
)

followup_plot_grid <- seq(
  0,
  20,
  by = 0.25
)

followup_axis_min <- 0
followup_axis_max <- max(followup_plot_grid)
```

### 4.2 Generate baseline-age anchored predictions

```{r}
# Inputs:
#   model_x = one fitted Class 2 model object for one test × sex stratum
#   data_x  = the analysis data used to fit that model
#
# Output:
#   A prediction grid with one row per baseline-age anchor × follow-up time,
#   including attained age and the model-predicted cognitive score.
make_class2_anchor_predictions <- function(
    model_x,
    data_x
) {

  #  Prediction grid
  #   - selected baseline-age anchors, e.g., 30, 35, ..., 85
  #   - follow-up times, e.g., 0, 0.25, ..., 20 years
  prediction_grid <- expand_grid(
    baseline_age_anchor = baseline_age_anchors,
    t_followup = followup_plot_grid
  ) |>
    mutate(
      # baseline_age is the model variable; baseline_age_anchor is kept for plotting.
      baseline_age = baseline_age_anchor,
       # Plot Class 2 follow-up trajectories on the attained-age scale.
      attained_age = baseline_age + t_followup,
      
      # Hold birth year fixed at the centered reference value.
      birth_year_c = 0,
      
      # Required grouping variables for predict(); random effects are excluded below.
      idtype_label = factor(
        "Offspring",
        levels = c("Original", "Offspring", "Gen3")
      ),
      dbgap_subject_id = data_x$dbgap_subject_id[1]
    ) |>
    filter(
      attained_age >= plot_age_min,
      attained_age <= plot_age_max
    )
  
# Fixed-effect predictions only; no subject- or cohort-specific random effects.
  prediction_grid$predicted_score <- as.numeric(
    predict(
      model_x,
      newdata = prediction_grid,
      re.form = NA,
      allow.new.levels = TRUE
    )
  )
  
  prediction_grid |>
    transmute(
      baseline_age_anchor,
      t_followup,
      attained_age,
      birth_year_c,
      predicted_score
    )
}
```



### 4.3 Apply prediction function to all Class 2 models

```{r}
class2_anchor_trajectory_predictions <- class2_fits_success |>
  mutate(
    anchor_predictions = map2(
      model,
      data,
      ~ make_class2_anchor_predictions( # defined in 4.2
        model_x = .x,
        data_x = .y
      )
    )
  ) |>
  select(
    test,
    sex_label,
    anchor_predictions
  ) |>
  unnest(
    anchor_predictions
  ) |>
  mutate(
    sex_label = factor(
      sex_label,
      levels = c("Women", "Men")
    )
  ) |>
  left_join(
    test_dictionary_sheet |>
      select(
        test,
        test_label,
        xprize_subdomain,
        range
      ),
    by = "test"
  ) |>
  arrange(
    xprize_subdomain,
    test,
    sex_label,
    baseline_age_anchor,
    t_followup
  )
```

```{r}
class2_anchor_trajectory_predictions |>
  filter(
    test == "lmi"
  ) |>
  summarise(
    n_rows = n(),
    min_baseline_age_anchor = min(baseline_age_anchor),
    max_baseline_age_anchor = max(baseline_age_anchor),
    min_attained_age = min(attained_age),
    max_attained_age = max(attained_age),
    min_predicted_score = min(predicted_score, na.rm = TRUE),
    max_predicted_score = max(predicted_score, na.rm = TRUE)
  ) |>
  knitr::kable(digits = 2)
```


### 4.4 Create curve-based Y-axis ranges

```{r}
class2_y_axis_table <- class2_anchor_trajectory_predictions |>
  group_by(test) |>
  summarise(
    y_axis = list(
      make_curve_y_axis( # defined in R/helpers.R
        curve_x = cur_data(),
        y_var = "predicted_score"
      )
    ),
    .groups = "drop"
  ) |>
  mutate(
    y_min = map_dbl(y_axis, ~ .x$limits[1]),
    y_max = map_dbl(y_axis, ~ .x$limits[2]),
    y_breaks = map(y_axis, ~ .x$breaks)
  ) |>
  select(
    test,
    y_min,
    y_max,
    y_breaks
  )

class2_y_axis_table |>
  knitr::kable(digits = 2)
```






### 4.5 Quick check trajectory counts for one test

```{r}
class2_anchor_trajectory_predictions |>
  filter(
    test == "lmi"
  ) |>
  count(
    sex_label,
    baseline_age_anchor
  ) |>
  arrange(
    sex_label,
    baseline_age_anchor
  ) |>
  knitr::kable()

```





### 4.6 Define Class 2 plotting function

```{r}
# Background points show baseline age × maximum follow-up on the right axis.
# Foreground curves show fixed-effect predicted scores by baseline-age anchor.

plot_class2_baseline_age_prediction_one_test <- function(
    background_data,
    prediction_data,
    test_dictionary,
    test_x,
    y_axis_table = NULL
) {
  
  # Get the display label for the current test.
  test_label_x <- test_dictionary |>
    filter(test == test_x) |>
    distinct(test_label) |>
    pull(test_label) |>
    first()
  
  # Use shared y-axis limits when supplied.
  if (is.null(y_axis_table)) {
    
    curve_x <- prediction_data |>
      filter(test == test_x)
    
    # make_curve_y_axis() is defined in R/helpers.R
    y_axis_x <- make_curve_y_axis(
      curve_x = curve_x,
      y_var = "predicted_score"
    )
    
    y_limits_x <- y_axis_x$limits
    y_breaks_x <- y_axis_x$breaks
    
  } else {
    
    y_axis_x <- y_axis_table |>
      filter(test == test_x) |>
      slice(1)
    
    y_limits_x <- c(
      y_axis_x$y_min,
      y_axis_x$y_max
    )
    
    y_breaks_x <- y_axis_x$y_breaks[[1]]
  }
  
  # Map maximum follow-up years onto the cognitive-score y-axis so the background support points can be drawn in the same panel.
  score_min <- y_limits_x[1]
  score_max <- y_limits_x[2]
  
  followup_to_score <- function(y) {
    score_min +
      (y - followup_axis_min) /
      (followup_axis_max - followup_axis_min) *
      (score_max - score_min)
  }
  
  # Inverse transform used for the right-side secondary axis.
  score_to_followup <- function(y) {
    followup_axis_min +
      (y - score_min) /
      (score_max - score_min) *
      (followup_axis_max - followup_axis_min)
  }
  
  ggplot() +
    # Background layer: Observed follow-up support
    geom_point(
      data = background_data |>
        filter(
          test == test_x
        ),
      aes(
        x = baseline_age,
        y = followup_to_score(max_followup),
        shape = idtype_label
      ),
      color = "grey55",
      alpha = 0.2,
      size = 0.8
    ) +
    
    # Foreground layer: Fixed-effect trajectories
    geom_line(
      data = prediction_data |>
        filter(
          test == test_x
        ),
      aes(
        x = attained_age,
        y = predicted_score,
        group = baseline_age_anchor,
        color = baseline_age_anchor
      ),
      linewidth = 0.95,
      alpha = 0.95
    ) +
    facet_wrap(
      vars(sex_label),
      nrow = 1
    ) +
    scale_color_viridis_c(
      name = "Baseline age",
      breaks = seq(
        min(baseline_age_anchors),
        max(baseline_age_anchors),
        by = 10
      ),
      limits = range(baseline_age_anchors)
    ) +
    scale_shape_manual(
      values = cohort_shapes,
      name = "Cohort"
    ) +
    scale_x_continuous(
      limits = c(plot_age_min, plot_age_max),
      breaks = seq(plot_age_min, plot_age_max, by = 10)
    ) +
    scale_y_continuous(
      name = "Cognitive score",
      breaks = y_breaks_x,
      sec.axis = sec_axis(
        trans = score_to_followup,
        name = "Maximum follow-up, years",
        breaks = seq(
          followup_axis_min,
          followup_axis_max,
          by = 5
        )
      )
    ) +
    coord_cartesian(
      ylim = y_limits_x
    ) +
    labs(
      title = paste0(
        "Class 2 · Follow-up as Time: ",
        test_label_x
      ),
      subtitle = "Background: maximum follow-up per subject (right axis); curves: fitted trajectories by baseline age",
      x = "Age at measurement"
    ) +
    guides(
      color = guide_colorbar(
        order = 1,
        title.position = "top",
        barwidth = unit(3.8, "cm"),
        barheight = unit(0.35, "cm")
      ),
      shape = guide_legend(
        order = 2,
        override.aes = list(
          color = "grey40",
          alpha = 0.7,
          size = 2
        )
      )
    ) +
    theme_bw(base_size = 12) +
    theme(
      legend.position = "bottom",
      legend.box = "horizontal",
      legend.title = element_text(size = 10.5),
      legend.text = element_text(size = 9.5),
      plot.title = element_text(size = 15, face = "bold"),
      plot.subtitle = element_text(size = 11.5, color = "grey30"),
      axis.title = element_text(size = 13),
      axis.text = element_text(size = 10.5),
      strip.background = element_rect(fill = "grey90"),
      strip.text = element_text(size = 11),
      panel.grid.minor = element_blank(),
      panel.spacing.x = unit(0.8, "lines")
    )
}
```



### 4.7 Check one Class 2 life-course figure
```{r}
plot_class2_baseline_age_prediction_one_test(
  background_data = baseline_followup_background,
  prediction_data = class2_anchor_trajectory_predictions,
  test_dictionary = test_dictionary_sheet,
  test_x = "lmd",
  y_axis_table = class2_y_axis_table
)
```

### 4.8 Save all Class 2 plots

```{r}
class2_plot_order <- class2_anchor_trajectory_predictions |>
  distinct(
    xprize_subdomain,
    test
  ) |>
  arrange(
    xprize_subdomain,
    test
  )

walk2(
  class2_plot_order$xprize_subdomain,
  class2_plot_order$test,
  \(subdomain_x, test_x) {
    
    plot_x <- plot_class2_baseline_age_prediction_one_test(
      background_data = baseline_followup_background,
      prediction_data = class2_anchor_trajectory_predictions,
      test_dictionary = test_dictionary_sheet,
      test_x = test_x,
      y_axis_table = class2_y_axis_table
    )
    
    subdomain_file_x <- subdomain_x |>
      str_to_lower() |>
      str_replace_all("&", "and") |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")
    
    test_file_x <- test_x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")
    
    ggsave(
      filename = file.path(
        class2_lifecourse_dir,
        paste0(
          "pooled3_class2_followup_as_time_",
          subdomain_file_x,
          "_",
          test_file_x,
          "_",
          output_suffix,
          ".pdf"
        )
      ),
      plot = plot_x,
      width = 12,
      height = 6
    )
  }
)
```


## 5. Output y-axis range
```{r}
class2_y_axis_table_path <- paste0(
  "../output/tables/class2_y_axis_table_for_lifecourse_plots", dir_tag, ".rds"
)

saveRDS(
  class2_y_axis_table,
  class2_y_axis_table_path
)
```