---
title: "Pooled 3-Cohort Neuropsych: Spontaneous Reversal (Class 2 Crude)"
author: "Minghui (Grace) Wang"
date: "`r Sys.Date()`"
output: html_document
---

<!--
Pipeline step: 05 (run order: 00 -> 01 -> 02 -> 03 -> 04 -> 05 -> 06 -> 07)
Purpose: Classify spontaneous reversal of cognitive aging using Class 2
         (follow-up-as-time) models. Equal analytic status with 06 (Class 1).
Inputs:  - Class 2 fitted model RDS from 01 (../output/models/)
         - neuropsych_pooled3_model_ready_xprize_with_covariates.csv (via setup_analysis_sample.R)
Key steps: Compute observed change (baseline to first follow-up); compute
           fixed-effect-only model changes at 5/10/15/20yr horizons; classify
           reversal per participant-test; build Table 4 (exclusive & cumulative);
           generate colored scatter plots.
Core functions:
          - compute_obs_change(), compute_model_change_class2(), classify_reversal(),
            make_table4(), plot_reversal_scatter(), higher_is_worse_tests
            from R/spontaneous_reversal_functions.R
          - make_pooled_test_dictionary(), write_table4_workbook()
            from R/workbook_writers.R
          - make_output_name()
            from R/helpers.R
Outputs: - Table 4 exclusive/cumulative workbooks (../output/tables/pooled3_05_.../)
         - Colored scatter PDFs (../output/plots/pooled3_05_.../)
Downstream: None (terminal output step for Class 2 reversal)

-->

## 0.Overview

This notebook finds **spontaneous reversal of cognitive aging** using Class 2 (follow-up-as-time) models.

**Algorithm.** We use each participant-test combination that has at least 2 valid visits for that test. We take the observed score change from baseline to the first follow-up and turn it into a per-year (annualized) change, and compare it to the model-estimated change over the 5, 10, 15, and 20-year horizons. Every model change is computed directly from the fitted model at the participant's exact baseline age (fixed effects only, `model.matrix %*% fixef`); we do not round the age to a grid.

For each horizon *h*, let `model_change` be the direction-adjusted model change, and `obs` be the direction-adjusted annualized observed change:

- If `model_change < 0` (expected decline): reversal when `obs > (-1) * model_change`.
- If `model_change > 0` (expected improvement): reversal when `obs > 2 * model_change`.
- If `model_change = 0` (no expected change): reversal when `obs > 0`.

For Trails tests (higher = worse), we flip the sign of both the observed change and the model change, so that positive always means improvement.

**Outputs.**

- **Table 4 (exclusive count)**: age-bin counts. Each participant-test combination is counted once, in the column for its maximum qualifying horizon.
- **Table 4 (cumulative count)**: age-bin counts. Each column counts everyone who qualifies at that horizon or higher.
- **Table 4 (cumulative percentage)**: the cumulative counts divided by N. This is the spontaneous reversal rate in each age bin.
- **Scatter plots**: annualized observed change vs baseline age, colored by the maximum reversal horizon. Restricted to baseline age 50 to 90, to match Table 4.

# 1. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 6
)

library(tidyverse)
library(lme4)
library(lmerTest)
library(openxlsx)

source("R/_load_all.R")

output_suffix <- format(Sys.Date(), "%m%d%Y")
```

## 1.1 Output folders

```{r choose_sample_and_dirs}
# Use the same 2-visit crude analysis sample as the fitted models.
SAMPLE <- "2visit"
covariates <- NULL
source("R/setup_analysis_sample.R")

model_tag <- if (SAMPLE == "2visit") "crude_2visit" else "crude"
dir_tag   <- if (SAMPLE == "2visit") "_2visit" else ""

class2_fits_file <- "pooled3_crude_2visit_class2_fits.rds"

plot_dir     <- paste0(
  "../output/plots/pooled3_05_spontaneous_reversal_class2_crude", dir_tag
)
table_dir    <- paste0(
  "../output/tables/pooled3_05_spontaneous_reversal_class2_crude", dir_tag
)
dir.create(plot_dir,     recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir,    recursive = TRUE, showWarnings = FALSE)
```

## 1.2 Load saved Class 2 models

```{r load_class2_models}
model_rds_dir <- "../output/models"

class2_fits <- readRDS(file.path(model_rds_dir, class2_fits_file))

class2_fits_success <- class2_fits |>
  filter(fit_success)

cat(
  "Loaded", nrow(class2_fits_success),
  "successfully fitted Class 2 model strata\n"
)

class2_fits_success |>
  select(test, sex_label, n_obs, n_subjects, subject_random_slope_used) |>
  knitr::kable()
```

## 1.3 Test dictionary and score direction

```{r test_dictionary}
test_dictionary <- make_pooled_test_dictionary(model_ready_pool) # defined in R/workbook_writers.R.
test_dictionary |> knitr::kable()
```

Trails A and Trails B are scored as completion time (higher = worse). For these tests, both observed and model changes are sign-flipped so that positive always means improvement.

```{r score_direction_qc}
# `higher_is_worse_tests` is defined in `R/spontaneous_reversal_functions.R`.

test_dictionary |>
  mutate(
    score_direction = if_else(test %in% higher_is_worse_tests, -1L, 1L),
    direction_note  = if_else(
      score_direction == -1L,
      "Higher is worse (completion time); sign-flipped",
      "Higher is better"
    )
  ) |>
  select(test, test_label, score_direction, direction_note) |>
  knitr::kable()
```

# 2. Participant-level reversal classification

# 2.1. Observed change (baseline to first follow-up)

We use each participant-test combination that has at least 2 valid visits for that test. The annualized change is `(score_followup1 - score_baseline) / followup_years`. The observed-change table (and the follow-up summary below) keeps only baseline age 50 to 90, to match Table 4 and the scatter plots.

## 2.1.1 Compute per stratum

```{r compute_obs_change}
# For each successfully fitted test × sex stratum, compute participant-level observed change from baseline to the first follow-up visit, then restrict to index baseline age 50-90.

# Input:
#   data = stratum-specific analysis data stored in class2_fits_success

# Output:
#   obs = a nested participant-level table with one row per participant-test combination (baseline age 50-90), including baseline age, follow-up interval, raw change, and annualized observed change.

class2_with_obs <- class2_fits_success |>
  mutate(
    obs = map(
      data,
      \(data_x) {
        data_x |>
          # defined in R/spontaneous_reversal_functions.R：
          compute_obs_change() |> 
          restrict_reversal_baseline_age()       
          # index baseline age 50-90 only, applied here so subject_obs, summaries, model changes, classification, Table 4, and plots all share the same sample.
      }
    )
  )

# Unnest stratum-specific observed-change tables into one participant-level table used for summaries and downstream reversal classification.
subject_obs <- class2_with_obs |>
  select(test, sex_label, obs) |>
  unnest(obs)

cat(
  "Observed change computed for",
  n_distinct(subject_obs$dbgap_subject_id),
  "unique subjects across",
  n_distinct(subject_obs$test), "tests\n"
)
```

## 2.1.2 Follow-up interval summary

```{r followup_summary}
subject_obs |>
  group_by(test, sex_label) |>
  summarise(
    n = n(),
    median_fu_years = median(followup_years),
    iqr_lo = quantile(followup_years, 0.25),
    iqr_hi = quantile(followup_years, 0.75),
    min_nonzero_fu_years = if_else(
      any(followup_years > 0, na.rm = TRUE),
      min(followup_years[followup_years > 0], na.rm = TRUE),
      NA_real_
    ),
    max_fu_years = max(followup_years),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1)
```

# 2.2. Model-estimated changes (Class 2)

We compute the model-derived threshold directly from the fitted model, using fixed effects only (`model.matrix %*% fixef`). We do not read any value from Table 2.

Class 2 separates baseline age from follow-up time. For a participant indexed at baseline age *A* and horizon *h*, the **reference start age** is *A − h*. The threshold is the expected *h*-year aging change from the reference start age *A − h* up to age *A*. In Class 2 this is a follow-up change that starts from reference baseline age *A − h* and runs *h* years (the reference-start-age contrast):

    Pred_Class2(baseline_age = A - h, t_followup = h)
    - Pred_Class2(baseline_age = A - h, t_followup = 0)

This uses the same contrast structure as Table 2 for an *h*-year follow-up change starting from baseline age *A − h*, but we evaluate it directly from the fitted model at the participant's exact age. We do not round to, or look up, the displayed 5-year Table 2 grid.

For example, if *A* = 72, the 5/10/15/20-year thresholds use:

- baseline_age = 67, `t_followup` 0 to 5
- baseline_age = 62, `t_followup` 0 to 10
- baseline_age = 57, `t_followup` 0 to 15
- baseline_age = 52, `t_followup` 0 to 20

Because we use the exact age, a 72.4-year-old uses reference start ages 67.4 / 62.4 / 57.4 / 52.4.

```{r compute_model_changes}

## Compute per stratum
class2_with_mc <- class2_with_obs |>
  mutate(
    model_changes = pmap(
      list(model, obs),
      function(model_x, obs_x) {
        # model_x = test × sex fitted Class 2 model
        # obs_x   = test × sex observed-change participant table
        compute_model_change_class2(model_x, unique(obs_x$baseline_age)) # defined in R/spontaneous_reversal_functions.R.
        # unique(obs_x$baseline_age) = participant-specific observed baseline age 
      }
    )
  )
```

# 2.3. Reversal classification

Join observed changes with model-estimated changes and apply the spontaneous reversal rule.

## 2.3.1 Classify

```{r classify_reversal}

# For each participant-test record, join observed change with exact-age model-estimated changes, then apply the reversal threshold rules.

reversal_subject_level <- class2_with_mc |>
  mutate(
    classified = pmap(
      list(obs, model_changes, test, sex_label),
      function(obs_x, mc_x, test_x, sex_x) {
        obs_x |>                                       # participant-level observed change
          left_join(mc_x, by = "baseline_age") |>      # exact-age model change
          mutate(test = test_x, sex_label = sex_x) |>
          classify_reversal()                          # defined in R/spontaneous_reversal_functions.R
      }
    )
  ) |>
  select(classified) |>
  unnest(classified) |>
  restrict_reversal_baseline_age()

cat(
  "Classified", nrow(reversal_subject_level),
  "participant-test combinations (baseline age 50-90) across",
  n_distinct(reversal_subject_level$test), "tests\n"
)
```

## 2.3.2 Reversal summary

```{r reversal_summary}
reversal_subject_level |>
  group_by(test, sex_label) |>
  summarise(
    n = n(),
    n_any_rev = sum(max_reversal_year > 0),
    pct_any = 100 * mean(max_reversal_year > 0),
    n_rev_5  = sum(reversal_5yr),
    n_rev_10 = sum(reversal_10yr),
    n_rev_15 = sum(reversal_15yr),
    n_rev_20 = sum(reversal_20yr),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1)
```


# 3. Output table 4

The participant-level classifications were summarized by baseline-age group in exclusive, cumulative, and cumulative-percentage formats.

## 3.1 Table 4 (exclusive form)

```{r table4_exclusive}
table4_exclusive <- make_table4(
  reversal_subject_level, test_dictionary, form = "exclusive"
)

# Preview one test.
table4_exclusive |>
  filter(test == "lmd") |>
  select(-xprize_subdomain, -test, -test_label, -age_bin) |>
  knitr::kable()
```

```{r table4_exclusive_save}
# Save CSV (analysis) + formatted XLSX (delivery). write_table4_workbook() and
# make_output_name() are defined in workbook_writers.R / helpers.R.
write_csv(
  table4_exclusive,
  file.path(
    table_dir,
    make_output_name(
      paste0("xprize_pooled3_", model_tag, "_class2_table4_exclusive"),
      "csv"
    )
  )
)

table4_excl_xlsx <- file.path(
  table_dir,
  make_output_name(
    paste0("xprize_pooled3_", model_tag, "_class2_table4_exclusive"),
    "xlsx"
  )
)
write_table4_workbook(table4_exclusive, table4_excl_xlsx, form = "exclusive")
```

## 3.2 Table 4 (cumulative form)

```{r table4_cumulative}
table4_cumulative <- make_table4(
  reversal_subject_level, test_dictionary, form = "cumulative"
)

# Preview one test.
table4_cumulative |>
  filter(test == "lmd") |>
  select(-xprize_subdomain, -test, -test_label, -age_bin) |>
  knitr::kable()
```

```{r table4_cumulative_save}
write_csv(
  table4_cumulative,
  file.path(
    table_dir,
    make_output_name(
      paste0("xprize_pooled3_", model_tag, "_class2_table4_cumulative"),
      "csv"
    )
  )
)

table4_cumul_xlsx <- file.path(
  table_dir,
  make_output_name(
    paste0("xprize_pooled3_", model_tag, "_class2_table4_cumulative"),
    "xlsx"
  )
)
write_table4_workbook(table4_cumulative, table4_cumul_xlsx, form = "cumulative")
```

## 3.3 Table 4 (cumulative percentage)

```{r table4_cumulative_pct}
table4_cumulative_pct <- make_table4(
  reversal_subject_level, test_dictionary, form = "cumulative_pct"
)

# Preview one test (values are proportions of N).
table4_cumulative_pct |>
  filter(test == "lmd") |>
  select(-xprize_subdomain, -test, -test_label, -age_bin) |>
  knitr::kable(digits = 3)
```

```{r table4_cumulative_pct_save}
write_csv(
  table4_cumulative_pct,
  file.path(
    table_dir,
    make_output_name(
      paste0("xprize_pooled3_", model_tag, "_class2_table4_cumulative_pct"),
      "csv"
    )
  )
)

table4_cumul_pct_xlsx <- file.path(
  table_dir,
  make_output_name(
    paste0("xprize_pooled3_", model_tag, "_class2_table4_cumulative_pct"),
    "xlsx"
  )
)
write_table4_workbook(
  table4_cumulative_pct, table4_cumul_pct_xlsx, form = "cumulative_pct"
)
```

# 4. Scatter plots

Table 4 gives only age-bin counts. These scatter plots show the **participant-level data behind those counts**, so we can see *who* is called a reverser and why.

Each point is one participant-test record:

- **x** = baseline age. **y** = the direction-adjusted annualized observed change (positive = improvement, for every test, after the Trails sign-flip). This is the exact number the reversal rule compares to the model threshold.
- The dashed line at **y = 0** separates observed decline (below) from observed improvement (above).
- **Color** = the maximum reversal horizon each record reached. Grey = none, and the color gets darker from 5y to 20y. A longer horizon means the point cleared a higher threshold, so it is a more extreme apparent reversal.
- Panels are split by sex. The age axis is 50 to 90, to match Table 4.

Scatter plots display the participant-level observations underlying Table 4. The x-axis shows baseline age and the y-axis shows direction-adjusted annualized change. Points are colored by maximum qualifying reversal horizon and panels are separated by sex.

## 4.1 Check one test

```{r scatter_check, fig.width=12, fig.height=6}
# defined in spontaneous_reversal_functions.R
plot_reversal_scatter(reversal_subject_level, test_dictionary, "lmd")
```

## 4.2 Save all tests

```{r scatter_save_all}
scatter_order <- reversal_subject_level |>
  distinct(test) |>
  left_join(
    test_dictionary |> select(test, xprize_subdomain),
    by = "test"
  ) |>
  arrange(xprize_subdomain, test)

walk2(
  scatter_order$xprize_subdomain,
  scatter_order$test,
  function(subdomain_x, test_x) {

    p <- plot_reversal_scatter(
      reversal_subject_level, test_dictionary, test_x
    )

    subdomain_file <- subdomain_x |>
      str_to_lower() |>
      str_replace_all("&", "and") |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")

    test_file <- test_x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")

    ggsave(
      filename = file.path(
        plot_dir,
        paste0(
          "class2_reversal_scatter_",
          subdomain_file, "_", test_file,
          "_", output_suffix, ".pdf"
        )
      ),
      plot = p,
      width = 12,
      height = 6
    )
  }
)

cat("Scatter plots saved to:", plot_dir, "\n")
```

