---
title: "Pooled 3-Cohort Neuropsych: Class 1 Life-Course Plots"
author: "Minghui (Grace) Wang"
date: "`r Sys.Date()`"
output: html_document
---

<!--
Pipeline step: 04 (run order: 00 -> 01 -> 02 -> 03 -> 04 -> 05 -> 06)
Purpose: Generate Class 1 (age-as-time) life-course plots with observed score
         background scatter and emmeans-based fitted curves.
         Requires 03 to have run first (reads Class 2 y-axis range for alignment).
Inputs:  - Class 1 fitted model RDS from 01 (../output/models/)
         - Class 2 y-axis range table RDS from 03 (../output/tables/)
         - neuropsych_pooled3_model_ready_xprize_with_covariates.csv (via setup_analysis_sample.R)
Key steps: Load Class 2 y-axis table for aligned scaling; generate emmeans-based
           prediction curves at ages 30–90 by 1; overlay observed scores colored
           by sex and shaped by cohort.
Core functions: make_curve_y_axis() (helpers.R), emmeans
Outputs: Class 1 life-course PDFs (../output/plots/pooled3_04_lifecourse_plots_class1_crude.../)

-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,   message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(lme4)
library(lmerTest)
library(emmeans)
library(broom.mixed)
library(ggsci)

source("R/_load_all.R")

output_suffix <- format(Sys.Date(), "%m%d%Y")
```

## 0. Overview
This script generates Class 1 life-course plots for the pooled 3-cohort neuropsychological models. It loads fitted model objects from RDS (saved by `01_fit_models.Rmd`) instead of refitting.

**Run `neuropsych_pooled3cohorts_03_plot_lifecourse_curves_model2_crude.Rmd` (notebook 03) first.** This script reads the y-axis range table saved by the Class 2 script so that both sets of plots share the same y-axis scale per test.

The Class 1 model uses age at measurement as the time scale. For each test, the figure overlays model-predicted cognitive score curves on a background scatter plot showing empirical data support.

### 0.1 Choose analysis sample

```{r}
SAMPLE <- "2visit"
```

```{r}
covariates <- NULL
source("R/setup_analysis_sample.R")
```

```{r}
model_tag <- if (SAMPLE == "2visit") "crude_2visit" else "crude"
dir_tag   <- if (SAMPLE == "2visit") "_2visit" else ""

plot_dir <- paste0(
  "../output/plots/pooled3_04_lifecourse_plots_class1_crude", dir_tag
)

dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Load fitted Class 1 model objects

```{r}
model_rds_dir <- "../output/models"

class1_fits <- readRDS(
  file.path(
    model_rds_dir,
    "pooled3_crude_2visit_class1_fits.rds"
  )
)
class1_fits_success <- class1_fits |>
  filter(fit_success)

cat(
  "Loaded", nrow(class1_fits_success),
  "successfully fitted Class 1 model strata\n"
)

class1_fits_success |>
  select(
    test,
    sex_label,
    n_obs,
    n_subjects,
    subject_random_slope_used
  ) |>
  knitr::kable()
```
## 2. Prepare test dictionary and Class 1 prediction grid

```{r}
test_dictionary_sheet <- make_pooled_test_dictionary( #  defined in R/workbook_writers.R
  model_ready_pool 
)

class1_plot_ages <- seq(
  30,
  90,
  by = 1
)

test_dictionary_sheet |>
  knitr::kable()
```


## 3. Create Class 1 prediction curves

```{r}

make_class1_prediction_curve <- function(model_x, data_x) {
  
  emm_x <- emmeans::emmeans(
    model_x,
    specs = ~ cage_65,
    at = list(
      cage_65 = class1_plot_ages - 65,
      birth_year_c = 0
    ),
    data = data_x,
    lmer.df = "asymptotic"
  )
  
  summary(emm_x) |>
    as_tibble() |>
    clean_names() |>
    mutate(
      age = cage_65 + 65
    ) |>
    select(
      age,
      predicted_score = emmean,
      se,
      everything()
    )
}
```


```{r}
class1_prediction_curves <- class1_fits_success |>
  mutate(
    prediction_curve = map2(
      model,
      data,
      ~ make_class1_prediction_curve(
        model_x = .x,
        data_x = .y
      )
    )
  ) |>
  select(
    test,
    sex_label,
    prediction_curve
  ) |>
  unnest(
    prediction_curve
  ) |>
  left_join(
    test_dictionary_sheet |>
      select(
        test,
        test_label,
        xprize_subdomain,
        range
      ),
    by = "test"
  ) |>
  arrange(
    xprize_subdomain,
    test,
    sex_label,
    age
  )

class1_prediction_curves |>
  filter(
    test == "lmi"
  ) |>
  head(20) |>
  knitr::kable(digits = 2)
```





## 4. Class 1 · Age as Time life-course figures


- The background layer shows observed cognitive scores by age at measurement, with cohort indicated by point shape and sex indicated by color. 
- The foreground layer shows the fitted Class 1 model fixed-effect curves for women and men.


### 4.1 Plot settings

```{r}
plot_age_min <- 30
plot_age_max <- 90

sex_colors <- c(
  "Women" = "#CC4C8A",
  "Men" = "#0E63AC"
)

cohort_shapes <- c(
  "Original" = 17,
  "Offspring" = 16,
  "Gen3" = 15
)

background_point_alpha <- 0.18
background_point_size <- 0.85

curve_linewidth <- 1.4


```



```{r}
# Load shared y-axis settings from codebook 03.

class2_y_axis_table <- readRDS(
  paste0(
    "../output/tables/class2_y_axis_table_for_lifecourse_plots", dir_tag, ".rds"
  )
)

class2_y_axis_table |>
  knitr::kable(digits = 2)

missing_axis_tests <- class1_prediction_curves |>
  distinct(test) |>
  anti_join(
    class2_y_axis_table |>
      distinct(test),
    by = "test"
  )

if (nrow(missing_axis_tests) > 0) {
  stop(
    "Missing Class 2 y-axis range for: ",
    paste(missing_axis_tests$test, collapse = ", ")
  )
}
```






### 4.2 Define plotting function

```{r}
plot_class1_age_support_one_test <- function(
    scatter_data,
    curve_data,
    test_dictionary,
    test_x,
    y_axis_table = NULL
) {
  
  scatter_x <- scatter_data |>
    filter(
      test == test_x,
      complete_for_model,
      age_at_measure >= plot_age_min,
      age_at_measure <= plot_age_max
    ) |>
    drop_na(
      age_at_measure,
      score,
      sex_label,
      idtype_label
    ) |>
    mutate(
      sex_label = factor(
        sex_label,
        levels = c("Women", "Men")
      ),
      idtype_label = factor(
        idtype_label,
        levels = c("Original", "Offspring", "Gen3")
      )
    )
  
  curve_x <- curve_data |>
    filter(
      test == test_x,
      age >= plot_age_min,
      age <= plot_age_max
    ) |>
    drop_na(
      age,
      predicted_score,
      sex_label
    ) |>
    mutate(
      sex_label = factor(
        sex_label,
        levels = c("Women", "Men")
      )
    )
  
  test_label_x <- test_dictionary |>
    filter(test == test_x) |>
    distinct(test_label) |>
    pull(test_label) |>
    first()

  
  if (is.null(y_axis_table)) {
    
    y_axis_x <- make_curve_y_axis(
      curve_x = curve_x,
      y_var = "predicted_score"
    )
    
    y_limits_x <- y_axis_x$limits
    y_breaks_x <- y_axis_x$breaks
    
  } else {
    
    y_axis_x <- y_axis_table |>
      filter(test == test_x) |>
      slice(1)
    
    y_limits_x <- c(
      y_axis_x$y_min,
      y_axis_x$y_max
    )
    
    y_breaks_x <- y_axis_x$y_breaks[[1]]
  }
  
  ggplot() +
    geom_point(
      data = scatter_x,
      aes(
        x = age_at_measure,
        y = score,
        color = sex_label,
        shape = idtype_label
      ),
      alpha = background_point_alpha,
      size = background_point_size
    ) +
    geom_ribbon(
      data = curve_x,
      aes(
        x = age,
        ymin = asymp_lcl,
        ymax = asymp_ucl,
        fill = sex_label
      ),
      alpha = 0.12,
      color = NA
    ) +
    geom_line(
      data = curve_x,
      aes(
        x = age,
        y = predicted_score,
        color = sex_label
      ),
      linewidth = curve_linewidth
    ) +
    scale_color_manual(
      values = sex_colors,
      name = "Sex"
    ) +
    scale_shape_manual(
      values = cohort_shapes,
      name = "Cohort"
    ) +
    scale_x_continuous(
      limits = c(plot_age_min, plot_age_max),
      breaks = seq(plot_age_min, plot_age_max, by = 10)
    ) +
    scale_y_continuous(
      name = "Cognitive score",
      breaks = y_breaks_x
    ) +
    coord_cartesian(
      ylim = y_limits_x
    ) +
    labs(
      title = paste0(
        "Class 1 · Age as Time: ",
        test_label_x
      ),
      subtitle = "Observed scores in background with fitted age-as-time curves by sex",
      x = "Age at measurement"
    ) +
    theme_bw(base_size = 12) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 15, face = "bold"),
      plot.subtitle = element_text(size = 11.5, color = "grey30"),
      axis.title = element_text(size = 13),
      axis.text = element_text(size = 10.5),
      panel.grid.minor = element_blank()
    )
}
```


```{r, fig.width=11, fig.height=6}
plot_class1_age_support_one_test(
  scatter_data = model_ready_analysis,
  curve_data = class1_prediction_curves,
  test_dictionary = test_dictionary_sheet,
  test_x = "lmd",
  y_axis_table = class2_y_axis_table
)
```



### 4.3 Save all Class 1 age-support figures

```{r}
class1_plot_test_order <- class1_prediction_curves |>
  distinct(
    xprize_subdomain,
    test
  ) |>
  arrange(
    xprize_subdomain,
    test
  )

walk2(
  class1_plot_test_order$xprize_subdomain,
  class1_plot_test_order$test,
  \(subdomain_x, test_x) {
    
    plot_x <- plot_class1_age_support_one_test(
      scatter_data = model_ready_analysis,
      curve_data = class1_prediction_curves,
      test_dictionary = test_dictionary_sheet,
      test_x = test_x,
      y_axis_table = class2_y_axis_table
    )
    
    subdomain_file_x <- subdomain_x |>
      str_to_lower() |>
      str_replace_all("&", "and") |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")
    
    test_file_x <- test_x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")
    
    ggsave(
      filename = file.path(
        plot_dir,
        paste0(
          "pooled3_class1_age_as_time_lifecourse_curve_",
          subdomain_file_x,
          "_",
          test_file_x,
          "_",
          output_suffix,
          ".pdf"
        )
      ),
      plot = plot_x,
      width = 11,
      height = 6
    )
  }
)
```

