---
title: "Pooled 3-Cohort Neuropsych: Formatted Output Tables (Class 1 & Class 2)"
author: "Minghui (Grace) Wang"
date: "`r Sys.Date()`"
output: html_document
---

This script loads the fitted model objects saved by `01_fit_models.Rmd` and produces the formatted Excel workbook containing Tables 1–3 for both model classes, plus summary sheets.

Table 1: predicted means at index ages.
Table 2: expected change over 5/10/15/20-year horizons.
Table 3: percentage decline calculated from Table 2 estimated changes and Table 1 estimated means.

<!--
Pipeline step: 02 (run order: 00 -> 01 -> 02 -> 03 -> 04 -> 05 -> 06)
Purpose: Generate formatted Excel workbooks with Tables 1–3 for both Class 1 and
         Class 2 models; includes test dictionary, data summary, and follow-up sheets.
Inputs:  - Class 1 and Class 2 fitted model RDS from 01 (../output/models/)
         - neuropsych_pooled3_model_ready_xprize_with_covariates.csv (via setup_analysis_sample.R)
Key steps: Build emmeans-based Table 1 (predicted means) and Table 2 (expected
           change over 5/10/15/20yr) for each model class; write formatted Excel
           workbook with per-test sheets and summary sheets.
Core functions: table_class1.R, table_class2.R, workbook_writers.R
Outputs: xprize_pooled3_{model_version}_formatted_tables_by_test.xlsx (../output/tables/)
Downstream: None (terminal output step for tables)
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)
library(lme4)
library(lmerTest)
library(emmeans)
library(openxlsx)

source("R/_load_all.R")

output_suffix <- format(Sys.Date(), "%m%d%Y")
```



## 0.1 Choose analysis sample

Must match the `SAMPLE` setting used when running `01_fit_models.Rmd`.

```{r}
SAMPLE <- "2visit"   # "full" or "2visit"
```

# 1. Load model-ready data and reconstruct analysis sample

The summary sheets in the workbook describe the analysis sample (subject counts, follow-up distributions). To build them we need the same data that `01_fit_models.Rmd` used for fitting.

```{r}
covariates <- NULL
source("R/setup_analysis_sample.R")
```

# 2. Load fitted model objects

```{r}
model_rds_dir <- "../output/models"

class1_fits <- read_rds(
  file.path(
    model_rds_dir,
    paste0("pooled3_", model_tag, "_class1_fits.rds")
  )
)

class2_fits <- read_rds(
  file.path(
    model_rds_dir,
    paste0("pooled3_", model_tag, "_class2_fits.rds")
  )
)

class1_fits_success <- class1_fits |>
  filter(fit_success)

class2_fits_success <- class2_fits |>
  filter(fit_success)

cat(
  "Class 1:", nrow(class1_fits_success), "strata;",
  "Class 2:", nrow(class2_fits_success), "strata\n"
)
```


# 3. Define output settings

```{r}
index_ages_table1 <- seq(
  50,
  90,
  by = 5
)

index_ages_table2 <- seq(
  30,
  85,
  by = 5
)

change_years <- c(
  5,
  10,
  15,
  20
)
```

# 4. Create table objects for Excel workbook

The following code shows how we generate Table 1-2. For Table 3, the `write_table3_block()` in `workbook_writers.R` converts each Table 2 estimated change into a percentage using the corresponding Table 1 estimated mean. In the current implementation, the denominator is the estimated mean at the end age (`age + years_of_change`).

## 4.1 Class 1 table objects

```{r}
class1_table1 <- class1_fits_success |>
  mutate(
    table1 = pmap(
      list(
        model,
        data,
        test,
        sex_label
      ),
      ~make_pooled_class1_table1(  # defined in R/table_class1.R
        model_x = ..1,
        data_x = ..2,
        test_x = ..3,
        sex_x = ..4
      )
    )
  ) |>
  select(
    model_version,
    model_class,
    n_obs,
    n_subjects,
    n_cohorts,
    subject_random_slope_used,
    random_effect_structure,
    table1
  ) |>
  unnest(table1) |>
  clean_names() |>
  arrange(
    test,
    sex_label,
    age
  )

class1_table1_wide <- format_pooled_class1_table1_wide(  # defined in R/table_class1.R
  class1_table1
)
```

```{r}
class1_table2 <- class1_fits_success |>
  mutate(
    table2 = pmap(
      list(
        model,
        data,
        test,
        sex_label
      ),
      ~ make_pooled_class1_table2(   # defined in R/table_class1.R
        model_x = ..1,
        data_x = ..2,
        test_x = ..3,
        sex_x = ..4
      )
    )
  ) |>
  select(
    model_version,
    model_class,
    n_obs,
    n_subjects,
    n_cohorts,
    subject_random_slope_used,
    random_effect_structure,
    table2
  ) |>
  unnest(table2) |>
  clean_names() |>
  arrange(
    test,
    sex_label,
    age,
    years_of_change
  )

class1_table2_wide <- format_pooled_class1_table2_wide(  # defined in R/table_class1.R
  class1_table2
)
```

## 4.2 Class 2 table objects

```{r}
class2_table1 <- class2_fits_success |>
  mutate(
    table1 = pmap(
      list(
        model,
        data,
        test,
        sex_label
      ),
      ~ make_pooled_class2_table1(  # defined in R/table_class2.R
        model_x = ..1,
        data_x = ..2,
        test_x = ..3,
        sex_x = ..4
      )
    )
  ) |>
  select(
    model_version,
    model_class,
    n_obs,
    n_subjects,
    n_cohorts,
    subject_random_slope_used,
    random_effect_structure,
    table1
  ) |>
  unnest(table1) |>
  clean_names() |>
  arrange(
    test,
    sex_label,
    baseline_age
  )

class2_table1_wide <- format_pooled_class2_table1_wide(  # defined in R/table_class2.R
  class2_table1
)
```

```{r}
class2_table2 <- class2_fits_success |>
  mutate(
    table2 = pmap(
      list(
        model,
        data,
        test,
        sex_label
      ),
      ~ make_pooled_class2_table2(  # defined in R/table_class2.R
        model_x = ..1,
        data_x = ..2,
        test_x = ..3,
        sex_x = ..4
      )
    )
  ) |>
  select(
    model_version,
    model_class,
    n_obs,
    n_subjects,
    n_cohorts,
    subject_random_slope_used,
    random_effect_structure,
    table2
  ) |>
  unnest(table2) |>
  clean_names() |>
  arrange(
    test,
    sex_label,
    baseline_age,
    years_of_change
  )

class2_table2_wide <- format_pooled_class2_table2_wide(  # defined in R/table_class2.R
  class2_table2
)
```



# 5. Workbook summary sheets

## 5.1 Test dictionary sheet

```{r}
test_dictionary_sheet <- make_pooled_test_dictionary(
  model_ready_analysis
)

test_dictionary_sheet |>
  knitr::kable()
```

## 5.2 Pooled data summary sheet

```{r}
summary_sheets <- make_pooled_workbook_summary_sheets(
  dat = model_ready_analysis,
  test_dictionary_sheet = test_dictionary_sheet,
  model_required_vars = model_required_vars
)
```

```{r}
data_summary_sheet <- summary_sheets$data_summary_sheet
data_summary_sheet |>
  knitr::kable(digits = 1) 
```

## 5.3 Pooled follow-up summary sheet

```{r}
followup_summary_sheet <- summary_sheets$followup_summary_sheet
followup_summary_sheet |>
  knitr::kable()
```

# 6. Write formatted Excel workbook

Each sheet corresponds to one neuropsychological test. The workbook also includes summary sheets for the test dictionary, pooled data summary, and pooled follow-up summary.

```{r}
formatted_pooled_path <- file.path(
  model_table_dir,
  make_output_name(
    paste0(
      "xprize_pooled3_",
      model_version,
      "_formatted_tables_by_test"
    ),
    "xlsx"
  )
)

write_all_pooled_test_workbook_with_summary( # is defined in R/workbook_writers.R
  dat = model_ready_analysis,
  class1_t1 = class1_table1_wide,
  class1_t2 = class1_table2_wide,
  class2_t1 = class2_table1_wide,
  class2_t2 = class2_table2_wide,
  test_dictionary_sheet = test_dictionary_sheet,
  data_summary_sheet = data_summary_sheet,
  followup_summary_sheet = followup_summary_sheet,
  out_file = formatted_pooled_path
)

```
