---
title: "Pooled 3-Cohort Neuropsych: Fit Class 1 and Class 2 Crude Models"
author: "Minghui (Grace) Wang"
date: "`r Sys.Date()`"
output: html_document
---

This script fits crude (without any covariates) pooled Class 1 and Class 2 mixed-effects models for neuropsych tests in the Framingham Study Original, Offspring, and Gen3 cohorts.

The fitted model objects are saved as RDS files and used by downstream scripts for formatted tables, life-course plots, and spontaneous reversal analyses.


<!--
Pipeline step: 01 (run order: 00 -> 01 -> 02 -> 03 -> 04 -> 05 -> 06)
Purpose: Fit Class 1 (age-as-time) and Class 2 (follow-up-as-time) mixed-effects
         models for each test × sex stratum using the 2-visit analysis sample.
Inputs:  neuropsych_pooled3_model_ready_xprize_with_covariates.csv (via setup_analysis_sample.R)
Key steps: Define Class 1/Class 2 formula text; check sample sizes; fit models
           with automatic fallback from random slope to intercept-only; save RDS.
Core functions: fit_one_pooled_model_class() (model_fitting.R),
                make_pooled_formula_text() (model_formulas.R)
Outputs: - pooled3_crude_2visit_class1_fits.rds (../output/models/)
         - pooled3_crude_2visit_class2_fits.rds (../output/models/)
-->


## 0.  Set up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)
library(lme4)
library(lmerTest)
library(emmeans)
library(broom.mixed)
library(openxlsx)

source("R/_load_all.R")

output_suffix <- format(Sys.Date(), "%m%d%Y")
```

The `SAMPLE` switch controls whether models are fit on the full pooled sample or the repeated-measures subset (subjects with ≥2 model-ready visits per test). Setting it once here tags every downstream output so the two runs never collide.

```{r}
SAMPLE <- "2visit"   # "full" or "2visit"
covariates <- NULL
source("R/setup_analysis_sample.R") # data import is finished in this step
```


## 1. Model specification

We fit the pooled 3-cohort models separately within each `test × sex` stratum.

The three Framingham cohorts are pooled into one model, but cohort structure is retained through cohort-level random effects. Subject-level random effects account for repeated measurements within the same participant. This crude model includes centered year of birth (`birth_year_c`) as a fixed-effect covariate. `birth_year_c = 0` represents the pooled 3-cohort median birth year.

### Class 1 model: age as the time scale

Class 1 models score as a function of age at measurement. Age is centered at 65 (`cage_65`) and modeled with both linear and quadratic terms.

```r
score ~ cage_65 + I(cage_65^2) + birth_year_c +
  (1 + cage_65 | idtype_label) +
  (1 + cage_65 | dbgap_subject_id)
```

Model terms:

- `score`: neuropsychological test score.
- `cage_65`: age at measurement centered at 65 years.
- `birth_year_c`: estimated birth year centered at the pooled 3-cohort median birth year.
- `idtype_label`: Framingham cohort label.
- `dbgap_subject_id`: participant ID.
- `(1 + cage_65 | idtype_label)`: cohort-level random intercept and age slope.
- `(1 + cage_65 | dbgap_subject_id)`: participant-level random intercept and age slope.


### Class 2 model: follow-up time as the time scale

Class 2 models score as a function of baseline age and follow-up time. The main model allows baseline age to modify both linear and quadratic follow-up change.

```r
score ~ baseline_age * (t_followup + I(t_followup^2)) + birth_year_c +
  (1 + t_followup | idtype_label) +
  (1 + t_followup | dbgap_subject_id)
```

Model terms:

- `score`: neuropsychological test score.
- `baseline_age`: age at first model-ready visit for that test.
- `t_followup`: years since baseline visit.
- `baseline_age * (t_followup + I(t_followup^2))`: baseline age main effect plus interactions with linear and quadratic follow-up time.
- `birth_year_c`: estimated birth year centered at the pooled 3-cohort median birth year.
- `idtype_label`: Framingham cohort label.
- `dbgap_subject_id`: participant ID.
- `(1 + t_followup | idtype_label)`: cohort-level random intercept and follow-up slope.
- `(1 + t_followup | dbgap_subject_id)`: participant-level random intercept and follow-up slope.

### The fallback models

The fallback model keeps the same fixed effects and cohort-level random effects, but uses a participant-level random intercept only.

For both Class 1 and Class 2, the primary model is fit first. If the participant-level random slope is not identifiable, the model is refit using the fallback random-effect structure.

The fallback structure is mainly relevant in the full-sample analysis, especially for test strata with many baseline-only participants. In the current analysis with 2+ visit sample (`SAMPLE = "2visit"`), all fitted models used the primary random-effect structure, so no fallback models were needed.

## 2. Define crude model settings

```{r}
# functions can be found in `model_formulas.R`

class1_primary_formula_text <- make_pooled_formula_text(
  model_class = "class1",
  covariates = covariates,
  subject_random_slope = TRUE
)

class1_fallback_formula_text <- make_pooled_formula_text(
  model_class = "class1",
  covariates = covariates,
  subject_random_slope = FALSE
)

class2_primary_formula_text <- make_pooled_formula_text(
  model_class = "class2",
  covariates = covariates,
  subject_random_slope = TRUE
)

class2_fallback_formula_text <- make_pooled_formula_text(
  model_class = "class2",
  covariates = covariates,
  subject_random_slope = FALSE
)

model_formula_check <- tibble(
  model_class = c("class1", "class1", "class2", "class2"),
  model_type = c("primary", "fallback", "primary", "fallback"),
  subject_random_slope = c(TRUE, FALSE, TRUE, FALSE),
  formula = c(
    class1_primary_formula_text,
    class1_fallback_formula_text,
    class2_primary_formula_text,
    class2_fallback_formula_text
  )
)

model_formula_check |>
  knitr::kable()
```

## 3. Check model sample size before fitting

This section checks the number of observations and subjects available for each test × sex stratum in the full pooled dataset, before any sample restriction. The `complete_for_model` flag is computed once here and reused downstream.

```{r}
model_sample_check_by_test_sex <- model_ready_pool |>
  group_by(
    test,
    sex_label
  ) |>
  summarise(
    n_obs_all = n(),
    n_subjects_all = n_distinct(dbgap_subject_id),

    n_obs_model_ready = sum(complete_for_model),
    n_subjects_model_ready = n_distinct(
      dbgap_subject_id[complete_for_model]
    ),

    n_missing_score = sum(is.na(score)),
    n_missing_cage_65 = sum(is.na(cage_65)),
    n_missing_baseline_age = sum(is.na(baseline_age)),
    n_missing_t_followup = sum(is.na(t_followup)),
    n_missing_birth_year_c = sum(is.na(birth_year_c)),

    .groups = "drop"
  ) |>
  mutate(
    pct_obs_kept = 100 * n_obs_model_ready / n_obs_all,
    pct_subjects_kept = 100 * n_subjects_model_ready / n_subjects_all
  ) |>
  arrange(
    test,
    sex_label
  )

model_sample_check_by_test_sex |>
  knitr::kable(digits = 2)
```

```{r}
write_csv(
  model_sample_check_by_test_sex,
  file.path(
    model_table_dir,
    make_output_name(
      paste0("models_", model_tag, "_sample_check_by_test_sex"),
      "csv"
    )
  )
)
```


## 4. Restrict sample (2-visit only) and create analysis grid

When `SAMPLE = "2visit"`, we keep only subjects with at least two model-ready visits per test. A subject-level random slope needs at least two measurements; single-visit subjects make the slope variance unidentifiable and cause the fallback strata seen in the full-sample run. The restriction is per test because a subject may have repeated measures on one test but not another.

When `SAMPLE = "full"`, this section is a no-op — `model_ready_analysis` is just `model_ready_pool`.

```{r}
# Summarize the analysis sample restriction already applied in setup_analysis_sample.R


if (SAMPLE == "2visit") {

  min_visits_per_test <- 2

  visits_per_subject_test <- model_ready_pool |>
    filter(complete_for_model) |>
    count(test, sex_label, dbgap_subject_id, name = "n_visits")

  restriction_summary <- visits_per_subject_test |>
    group_by(test, sex_label) |>
    summarise(
      n_subjects_full = n(),
      n_subjects_repeated = sum(n_visits >= min_visits_per_test),
      n_obs_full = sum(n_visits),
      n_obs_repeated = sum(n_visits[n_visits >= min_visits_per_test]),
      .groups = "drop"
    ) |>
    mutate(
      n_subjects_dropped = n_subjects_full - n_subjects_repeated,
      pct_subjects_retained = 100 * n_subjects_repeated / n_subjects_full
    ) |>
    arrange(test, sex_label)

} else {
  restriction_summary <- tibble(
    note = "SAMPLE = 'full'; no 2-visit restriction was applied."
  )
}

restriction_summary |>
  knitr::kable(digits = 1)
```

```{r}
write_csv(
  restriction_summary,
  file.path(
    model_table_dir,
    make_output_name(
      paste0("models_", model_tag, "_restriction_summary"),
      "csv"
    )
  )
)
```


```{r}

analysis_grid <- model_ready_analysis |>
  # `complete_for_model` is created in `setup_analysis_sample.R`. It keeps rows with non-missing model-required variables.
  filter(complete_for_model) |>  
  group_by(
    test,
    sex_label
  ) |>
  nest() |>
  mutate(
    model_version = model_version,
    n_obs = map_int(data, nrow),
    n_subjects = map_int(
      data,
      ~ n_distinct(.x$dbgap_subject_id)
    ),
    n_cohorts = map_int(
      data,
      ~ n_distinct(.x$idtype)
    )
  ) |>
  ungroup() |>
  arrange(
    test,
    sex_label
  )

analysis_grid |>
  select(
    test,
    sex_label,
    n_obs,
    n_subjects,
    n_cohorts
  ) |>
  knitr::kable()
```

## 5. Fit crude pooled Class 1 and Class 2 models

This section fits the crude pooled Class 1 and Class 2 models separately for each test × sex stratum. For each model class, the primary model includes subject-level random intercepts and random slopes. If the primary model is not identifiable, the same fixed-effect model is refit using subject-level random intercepts only. Cohort-level random intercepts and random slopes are retained in both specifications.


```{r}

#`fit_one_pooled_model_class()` is saved in `R/model_fitting.R`

class1_fits <- fit_one_pooled_model_class(
  grid_x = analysis_grid,
  model_class_x = "class1"
)

class2_fits <-fit_one_pooled_model_class(
  grid_x = analysis_grid,
  model_class_x = "class2"
)

model_fits <- bind_rows(
  class1_fits,
  class2_fits
)
```

### 5.1 Check fallback model strata

This section lists the test × sex strata where the primary subject-level random slope model was not identifiable and the fallback subject-level random-intercept-only model was used.

```{r}
fallback_model_strata <- model_fits |>
  filter(
    fit_success,
    !subject_random_slope_used
  ) |>
  select(
    model_class,
    test,
    sex_label,
    n_obs,
    n_subjects,
    n_cohorts,
    random_effect_structure,
    primary_error_message
  ) |>
  arrange(
    model_class,
    test,
    sex_label
  )

fallback_model_strata |>
  knitr::kable()

```

```{r}
write_csv(
  fallback_model_strata,
  file.path(
    model_table_dir,
    make_output_name(
      paste0("models_", model_tag, "_fallback_model_strata"),
      "csv"
    )
  )
)
```

### 5.2 Check failed model strata
This section checks whether any test × sex models failed even after applying the fallback random-effect structure.

```{r}
failed_model_strata <- model_fits |>
  filter(
    !fit_success
  ) |>
  select(
    model_class,
    test,
    sex_label,
    n_obs,
    n_subjects,
    n_cohorts,
    primary_error_message,
    fallback_error_message
  ) |>
  arrange(
    model_class,
    test,
    sex_label
  )

failed_model_strata |>
  knitr::kable()

```

```{r}
write_csv(
  failed_model_strata,
  file.path(
    model_table_dir,
    make_output_name(
      paste0("models_", model_tag, "_failed_model_strata"),
      "csv"
    )
  )
)
```

### 5.3 Save fitted model objects

This section saves the fitted Class 1 and Class 2 model objects as RDS files. These saved objects are used by downstream scripts (02 tables, 03 and 04 life-course plots, 05 and 06 spontaneous reversal) so that the models do not need to be refit.

```{r}
model_rds_dir <- "../output/models"

dir.create(
  model_rds_dir,
  recursive = TRUE,
  showWarnings = FALSE
)

saveRDS(
  class1_fits,
  file.path(
    model_rds_dir,
    paste0("pooled3_", model_tag, "_class1_fits.rds")
  )
)

saveRDS(
  class2_fits,
  file.path(
    model_rds_dir,
    paste0("pooled3_", model_tag, "_class2_fits.rds")
  )
)

cat(
  "Saved class1_fits:",
  nrow(class1_fits), "strata\n",
  "Saved class2_fits:",
  nrow(class2_fits), "strata\n"
)
```
