---
title: "Pooled 3-Cohort Neuropsych: Spontaneous Reversal (Class 1 Crude)"
author: "Minghui (Grace) Wang"
date: "`r Sys.Date()`"
output: html_document
---

<!--
Notebook 06 applies the fitted Class 1 models to classify spontaneous reversal and produce the Class 1 Table 4 outputs and scatter plots.
The classification procedure parallels notebook 05; only the model-estimated age contrast differs.
-->

# 0. Overview

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 6
)

library(tidyverse)
library(janitor)
library(lme4)
library(lmerTest)
library(openxlsx)

source("R/_load_all.R")

output_suffix <- format(Sys.Date(), "%m%d%Y")
```

## Method

For each test and sex stratum, observed change was calculated from baseline to the first follow-up and annualized by the elapsed follow-up time.

Class 1 thresholds were evaluated at each participant's exact index age using fixed-effect predictions. For index age *A* and horizon *h*, the model contrast was:

    Pred(age = A) - Pred(age = A - h)

For each horizon:

- expected decline: threshold = `-model_change`;
- expected improvement: threshold = `2 × model_change`;
- no expected change: threshold = 0.

Observed and model-estimated changes were sign-reversed for Trails A and Trails B so that positive values consistently indicate improvement.

Outputs include exclusive counts, cumulative counts, cumulative percentages, and participant-level scatter plots for baseline ages 50–90.


# 1. Setup

## 1.1 Output folders

```{r choose_sample_and_dirs}
SAMPLE <- "2visit"
covariates <- NULL
source("R/setup_analysis_sample.R")

model_tag <- if (SAMPLE == "2visit") "crude_2visit" else "crude"
dir_tag   <- if (SAMPLE == "2visit") "_2visit" else ""

class1_fits_file <- paste0(
  "pooled3_",
  model_tag,
  "_class1_fits.rds"
)

plot_dir     <- paste0(
  "../output/plots/pooled3_06_spontaneous_reversal_class1_crude", dir_tag
)
table_dir    <- paste0(
  "../output/tables/pooled3_06_spontaneous_reversal_class1_crude", dir_tag
)
dir.create(plot_dir,     recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir,    recursive = TRUE, showWarnings = FALSE)
```

## 1.2 Load saved Class 1 models

```{r load_class1_models}
model_rds_dir <- "../output/models"

class1_fits <- readRDS(file.path(model_rds_dir, class1_fits_file))

class1_fits_success <- class1_fits |>
  filter(fit_success)

cat(
  "Loaded", nrow(class1_fits_success),
  "successfully fitted Class 1 model strata\n"
)

class1_fits_success |>
  select(test, sex_label, n_obs, n_subjects, subject_random_slope_used) |>
  knitr::kable()
```

## 1.3 Test dictionary

```{r test_dictionary}
test_dictionary <- make_pooled_test_dictionary(model_ready_pool)
test_dictionary |> knitr::kable()
```

## 1.4 Score direction

Trails A and Trails B are scored as completion time (higher = worse). For
these tests, both observed and model changes are sign-flipped so that
positive always means improvement.

```{r score_direction_qc}
test_dictionary |>
  mutate(
    score_direction = if_else(test %in% higher_is_worse_tests, -1L, 1L),
    direction_note  = if_else(
      score_direction == -1L,
      "Higher is worse (completion time); sign-flipped",
      "Higher is better"
    )
  ) |>
  select(test, test_label, score_direction, direction_note) |>
  knitr::kable()
```

# 2. Observed change (baseline to first follow-up)

Observed change was calculated from baseline to the first follow-up and annualized by the elapsed follow-up time. Records were restricted to index baseline ages 50–90 before calculating the model-based thresholds.

## 2.1 Compute per stratum

```{r compute_obs_change}
class1_with_obs <- class1_fits_success |>
  mutate(
    obs = map(
      data,
      \(data_x) {
        data_x |>
          compute_obs_change() |>                # defined in R/spontaneous_reversal_functions.R.
          restrict_reversal_baseline_age()       
      }
    )
  )

subject_obs <- class1_with_obs |>
  select(test, sex_label, obs) |>
  unnest(obs)

cat(
  "Observed change computed for",
  n_distinct(subject_obs$dbgap_subject_id),
  "unique subjects across",
  n_distinct(subject_obs$test), "tests\n"
)
```

## 2.2 Follow-up intervals

```{r followup_summary}
subject_obs |>
  group_by(test, sex_label) |>
  summarise(
    n = n(),
    median_fu_years = median(followup_years),
    iqr_lo = quantile(followup_years, 0.25),
    iqr_hi = quantile(followup_years, 0.75),
    max_fu_years = max(followup_years),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1)
```

# 3. Model-estimated changes (Class 1, exact age)

Class 1 model changes were evaluated directly from the fitted fixed effects at each participant's exact index age. The backward-age contrast follows the definition in the Method section.

## 3.1 Compute per stratum

```{r compute_model_changes}
class1_with_mc <- class1_with_obs |>
  mutate(
    model_changes = pmap(
      list(model, obs),
      function(model_x, obs_x) {
        compute_model_change_class1(model_x, unique(obs_x$baseline_age))
      }
    )
  )
```

# 4. Reversal classification

Observed changes were joined to the exact-age Class 1 contrasts before applying the horizon-specific reversal thresholds.

## 4.1 Classify

```{r classify_reversal}
reversal_subject_level <- class1_with_mc |>
  mutate(
    classified = pmap(
      list(obs, model_changes, test, sex_label),
      function(obs_x, mc_x, test_x, sex_x) {
        obs_x |>
          left_join(mc_x, by = "baseline_age") |>
          mutate(test = test_x, sex_label = sex_x) |>
          classify_reversal()
      }
    )
  ) |>
  select(classified) |>
  unnest(classified) |>
  restrict_reversal_baseline_age()

cat(
  "Classified", nrow(reversal_subject_level),
  "participant-test combinations (baseline age 50-90) across",
  n_distinct(reversal_subject_level$test), "tests\n"
)
```

## 4.2 Reversal summary

```{r reversal_summary}
reversal_subject_level |>
  group_by(test, sex_label) |>
  summarise(
    n = n(),
    n_any_rev = sum(max_reversal_year > 0),
    pct_any = 100 * mean(max_reversal_year > 0),
    n_rev_5  = sum(reversal_5yr),
    n_rev_10 = sum(reversal_10yr),
    n_rev_15 = sum(reversal_15yr),
    n_rev_20 = sum(reversal_20yr),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1)
```

# 5. Table 4 outputs

Participant-level spontaneous reversal classifications were summarized by baseline-age group in three formats:

- **Exclusive count:** maximum qualifying horizon;
- **Cumulative count:** maximum qualifying horizon at or above each column;
- **Cumulative percentage:** cumulative count divided by N.

## 5.1 Table 4 (exclusive form)

```{r table4_exclusive}
table4_exclusive <- make_table4(
  reversal_subject_level, test_dictionary, form = "exclusive"
)

table4_exclusive |>
  filter(test == "lmd") |>
  select(-xprize_subdomain, -test, -test_label, -age_bin) |>
  knitr::kable()
```

```{r table4_exclusive_save}

# write_table4_workbook() is defined in workbook_writers.R
# make_output_name() is defined in helpers.R.

write_csv(
  table4_exclusive,
  file.path(
    table_dir,
    make_output_name(
      paste0("xprize_pooled3_", model_tag, "_class1_table4_exclusive"),
      "csv"
    )
  )
)

table4_excl_xlsx <- file.path(
  table_dir,
  make_output_name(
    paste0("xprize_pooled3_", model_tag, "_class1_table4_exclusive"),
    "xlsx"
  )
)
write_table4_workbook(table4_exclusive, table4_excl_xlsx, form = "exclusive")
```

## 5.2 Table 4 (cumulative form)

```{r table4_cumulative}
table4_cumulative <- make_table4(
  reversal_subject_level, test_dictionary, form = "cumulative"
)

# Preview one test.
table4_cumulative |>
  filter(test == "lmd") |>
  select(-xprize_subdomain, -test, -test_label, -age_bin) |>
  knitr::kable()
```


```{r table4_cumulative_save}
write_csv(
  table4_cumulative,
  file.path(
    table_dir,
    make_output_name(
      paste0("xprize_pooled3_", model_tag, "_class1_table4_cumulative"),
      "csv"
    )
  )
)

table4_cumul_xlsx <- file.path(
  table_dir,
  make_output_name(
    paste0("xprize_pooled3_", model_tag, "_class1_table4_cumulative"),
    "xlsx"
  )
)
write_table4_workbook(table4_cumulative, table4_cumul_xlsx, form = "cumulative")
```

## 5.3 Table 4 (cumulative percentage)

```{r table4_cumulative_pct}
table4_cumulative_pct <- make_table4(
  reversal_subject_level, test_dictionary, form = "cumulative_pct"
)

# Preview one test (values are proportions of N).
table4_cumulative_pct |>
  filter(test == "lmd") |>
  select(-xprize_subdomain, -test, -test_label, -age_bin) |>
  knitr::kable(digits = 3)
```

```{r table4_cumulative_pct_save}
write_csv(
  table4_cumulative_pct,
  file.path(
    table_dir,
    make_output_name(
      paste0("xprize_pooled3_", model_tag, "_class1_table4_cumulative_pct"),
      "csv"
    )
  )
)

table4_cumul_pct_xlsx <- file.path(
  table_dir,
  make_output_name(
    paste0("xprize_pooled3_", model_tag, "_class1_table4_cumulative_pct"),
    "xlsx"
  )
)
write_table4_workbook(
  table4_cumulative_pct, table4_cumul_pct_xlsx, form = "cumulative_pct"
)
```

# 6. Scatter plots

Scatter plots display the observations underlying Table 4. The x-axis shows index baseline age, and the y-axis shows direction-adjusted annualized observed change. Points are colored by maximum qualifying reversal horizon and panels are separated by sex.


## 6.1 Check one test

```{r scatter_check, fig.width=12, fig.height=6}
plot_reversal_scatter(reversal_subject_level, test_dictionary, "lmd")
```

## 6.2 Save all tests

```{r scatter_save_all}
scatter_order <- reversal_subject_level |>
  distinct(test) |>
  left_join(
    test_dictionary |> select(test, xprize_subdomain),
    by = "test"
  ) |>
  arrange(xprize_subdomain, test)

walk2(
  scatter_order$xprize_subdomain,
  scatter_order$test,
  function(subdomain_x, test_x) {

    p <- plot_reversal_scatter(
      reversal_subject_level, test_dictionary, test_x
    )

    subdomain_file <- subdomain_x |>
      str_to_lower() |>
      str_replace_all("&", "and") |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")

    test_file <- test_x |>
      str_to_lower() |>
      str_replace_all("[^a-z0-9]+", "_") |>
      str_replace_all("_+", "_") |>
      str_replace_all("^_|_$", "")

    ggsave(
      filename = file.path(
        plot_dir,
        paste0(
          "class1_reversal_scatter_",
          subdomain_file, "_", test_file,
          "_", output_suffix, ".pdf"
        )
      ),
      plot = p,
      width = 12,
      height = 6
    )
  }
)

cat("Scatter plots saved to:", plot_dir, "\n")
```



